Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-08 paper/paper.tex
revision of section 6.2

\subsection{Bounding the MIS enumeration space}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. In these graphs, a circle is a tensor, an arrow specifies execution direction of a function and $\circ$ is the Hadamard (element-wise) multiplication. (a) is the forward pass with algebra \Eq{eq:tropical} for computing $\alpha(G)$.
     (b) is the backward pass for computing boolean gradients as masks.
     (c) is the masked forward pass with algebra \Eq{eq:countingtropicalset} for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

When we use the algebra in \Eq{eq:countingtropicalset} to enumerate all MIS configurations, we find that the program stores significantly more intermediate configurations than necessary and thus incur significant overheads in space.
To speed up the computation and reduce space overhead, we bound the searching space using the MIS size $\alpha(G)$.
As shown in \Fig{fig:bounding}, (a) we first compute the value of $\alpha(G)$ with tropical algebra and cache all intermediate tensors.
(b) Then, we compute a boolean mask for each cached tensor, where we use a boolean true to represent a tensor element having a contribution to the MIS (i.e.\ with a non-zero gradient) and boolean false otherwise.
(c) Finally, we perform masked tensor network contraction using the element type with the algebra in \Eq{eq:countingtropicalset} to obtaining all configurations.
Note that these masks in fact correspond to tensor elements with non-zero gradients with respect to the MIS size; we compute these masks by back propagating the gradients.
To derive the back-propagation rule for tropical tensor contraction,
we first reduce the problem to finding the back-propagation rule of a tropical matrix multiplication $C = A B$.
Since $ C_{ik} = \bigoplus_{j} \ A_{ij} \odot B_{jk} = \max_{j} \ A_{ij} \odot B_{jk}$ with tropical algebra, we have the following inequality
\begin{equation}
    A_{ij} \odot B_{jk} \leq C_{ik}.
\end{equation}
Here $\leq$ on tropical numbers are the same as the real-number algebra.
The equality holds for some $j'$, which means $A_{ij'}$ and $B_{j'k}$ have contributions to $C_{ik}$.
Intuitively, one can use this relation to identify elements with nonzero gradients in $A$ and $B$ directly,
however, doing this might make us lost the advantage of using BLAS libraries~\cite{TropicalGEMM}.
Since $A_{ij} \odot B_{jk} = A_{ij} + B_{jk}$, one can move $B_{jk}$ to the right hand side of the inequality: 
\begin{equation}
    A_{ij} \leq C_{ik} \odot B_{jk}^{\circ -1}
\end{equation}
where ${}^{\circ -1}$ is the element-wise multiplicative inverse on tropical algebra (which is the additive inverse on real numbers).
The inequality still holds if we take the minimum over $k$: 
\begin{equation}
    A_{ij} \leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = \left(\max_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left(\bigoplus_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left( C^{\circ-1} B^{\mathsf{T}} \right)^{\circ -1}_{ij}.
\end{equation}
On the right hand side, we transform the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines~\cite{TropicalGEMM}.
Again, the equality holds if and only if the element $A_{ij}$ has a contribution to $C$ (i.e.\ having a non-zero gradient).
Let the gradient mask for $C$ be $\overline C$; the back-propagation rule for gradient masks reads
\begin{equation}\label{eq:adrule}
\overline{A}_{ij} = \delta \left(A_{ij}, \left( \left( C^{\circ-1} \circ \overline C \right) B^{\mathsf{T}} \right)_{ij}^{\circ -1} \right),
\end{equation}
where $\delta$ is the Dirac delta function that returns one if two arguments have the same value and zero otherwise, $\circ$ is the element-wise product, boolean false is treated as the tropical number $\mymathbb{0}$, and boolean true is treated as the tropical number $\mymathbb{1}$.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^{\mathsf{T}}$ by a tensor contraction.
With the above method, one can significantly reduce the space needed to store the intermediate configurations by setting the tensor elements masked false to zero during contraction.

>>>>

\subsection{Bounding the MIS enumeration space}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. In these graphs, a circle is a tensor, an arrow specifies execution direction of a function and $\circ$ is the Hadamard (element-wise) multiplication. $\overline A$ means the boolean mask of $A$. (a) is the forward pass with algebra \Eq{eq:tropical} for computing $\alpha(G)$.
     (b) is the backward pass for computing boolean gradients as masks.
     (c) is the masked forward pass with algebra \Eq{eq:countingtropicalset} for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

When we use the algebra in \Eq{eq:countingtropicalset} to enumerate all MIS configurations, we find that the program stores significantly more intermediate configurations than necessary and thus incur significant overheads in space.
To speed up the computation and reduce space overhead, we bound the searching space using the information from the computation of the MIS size $\alpha(G)$.
As shown in \Fig{fig:bounding}, (a) we first compute the value of $\alpha(G)$ with tropical algebra and cache all intermediate tensors.
(b) Then, we compute a boolean mask for each cached tensor, where we use a boolean true to represent a tensor element having a contribution to the MIS (i.e.\ with a non-zero gradient) and boolean false otherwise.
(c) Finally, we perform masked tensor network contraction (i.e.\ discarding the unnecessary intermediate configurations) using the element type with the algebra in \Eq{eq:countingtropicalset} to obtain all MIS configurations.
Note that these masks in fact correspond to tensor elements with non-zero gradients with respect to the MIS size; we compute these masks by back propagating the gradients.
To derive the back-propagation rule for tropical tensor contraction,
we first reduce the problem to finding the back-propagation rule of a tropical matrix multiplication $C = A B$.
Since $ C_{ik} = \bigoplus_{j} \ A_{ij} \odot B_{jk} = \max_{j} \ A_{ij} \odot B_{jk}$ with tropical algebra, we have the following inequality
\begin{equation}
    A_{ij} \odot B_{jk} \leq C_{ik}.
\end{equation}
Here $\leq$ on tropical numbers are the same as the real-number algebra.
The equality holds for some $j'$, which means $A_{ij'}$ and $B_{j'k}$ have contributions to $C_{ik}$.
Intuitively, one can use this relation to identify elements with nonzero gradients in $A$ and $B$,
but if doing this directly, one loses the advantage of using BLAS libraries~\cite{TropicalGEMM} for high performance.
Since $A_{ij} \odot B_{jk} = A_{ij} + B_{jk}$, one can move $B_{jk}$ to the right hand side of the inequality: 
\begin{equation}
    A_{ij} \leq C_{ik} \odot B_{jk}^{\circ -1}
\end{equation}
where ${}^{\circ -1}$ is the element-wise multiplicative inverse on tropical algebra (which is the additive inverse on real numbers).
The inequality still holds if we take the minimum over $k$: 
\begin{equation}
    A_{ij} \leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = \left(\max_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left(\bigoplus_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left( C^{\circ-1} B^{\mathsf{T}} \right)^{\circ -1}_{ij}.
\end{equation}
On the right hand side, we transform the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines~\cite{TropicalGEMM}.
Again, the equality holds if and only if the element $A_{ij}$ has a contribution to $C$ (i.e.\ having a non-zero gradient).
Let the gradient mask for $C$ be $\overline C$; the back-propagation rule for gradient masks reads
\begin{equation}\label{eq:adrule}
\overline{A}_{ij} = \delta \left(A_{ij}, \left( \left( C^{\circ-1} \circ \overline C \right) B^{\mathsf{T}} \right)_{ij}^{\circ -1} \right),
\end{equation}
where $\delta$ is the Dirac delta function that returns one if two arguments have the same value and zero otherwise, $\circ$ is the element-wise product, boolean false is treated as the tropical number $\mymathbb{0}$, and boolean true is treated as the tropical number $\mymathbb{1}$.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^{\mathsf{T}}$ by a tensor contraction.
With the above method, one can significantly reduce the space needed to store the intermediate configurations by setting the tensor elements masked false to zero during contraction.
